---
title: "Using the ExtendedLaplace Package"
author: "David Saah & Tomasz Kozubowski"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Using the ExtendedLaplace Package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#"
)
```

## Introduction

The `ExtendedLaplace` package provides tools for working with the Extended Laplace (EL) distribution, a generalization of the classical Laplace distribution. This distribution is characterized by four parameters: location $\mu$, scale \(\sigma > 0\), and a uniform noise range \(\delta > 0\).

The EL distribution arises as the sum \(Y = X + U\) where \(X \sim \text{Laplace}(\mu, \sigma)\) and \(U \sim \text{Uniform}(-\delta, \delta)\).

## Installation

To install the development version of this package from GitHub:

```r
# install.packages("devtools")
devtools::install_github("saahdavid/ExtendedLaplace")
```

## Functions

The package provides the following main functions:

- `dEL(y, mu, sigma, delta)`: Probability density function
- `pEL(y, mu, sigma, delta)`: Cumulative distribution function
- `qEL(u, mu, sigma, delta)`: Quantile function (inverse CDF)
- `rEL(n, mu, sigma, delta)`: Random number generation

## Examples

### Density
```{r}
library(ExtendedLaplace)
curve(dEL(x, mu = 0, sigma = 1, delta = 1), from = -5, to = 5, ylab = "Density", xlab = 'y')
```

### Distribution Function
```{r}
curve(pEL(x, mu = 0, sigma = 1, delta = 1), from = -5, to = 5, ylab = "CDF", xlab = 'y')
```

### Quantiles
```{r}
qEL(c(0.025, 0.5, 0.975), mu = 0, sigma = 1, delta = 1)
```

### Simulation
```{r}
samples <- rEL(10000, mu = 0, sigma = 1, delta = 1)
hist(samples, probability = TRUE, breaks = 40, main = "Simulated EL Data", xlab = 'y')
curve(dEL(x, mu = 0, sigma = 1, delta = 1), add = TRUE, col = "navy", lwd = 2)
```

## Theoretical Notes

The Extended Laplace distribution has the following form:

### PDF

$$
\begin{aligned}
g(y) = \frac{1}{4\delta} 
\begin{cases}
  e^{\frac{y- \mu + \delta}{\sigma}} - e^{\frac{y- \mu - \delta}{\sigma}}, & y < \mu - \delta \\
  2 - e^{-\frac{y - \mu + \delta}{\sigma}} - e^{\frac{y - \mu - \delta}{\sigma}}, & \mu - \delta \leq y < \mu + \delta \\
  e^{-\frac{y - \mu - \delta}{\sigma}} - e^{-\frac{y - \mu + \delta}{\sigma}}, & y \geq \mu + \delta
\end{cases}
\end{aligned}
$$

### CDF

$$
\begin{aligned}
    G(y) = \frac{1}{4\delta}
    \begin{cases}
        \sigma e^{\frac{y- \mu + \delta}{\sigma}} -  \sigma e^{\frac{y- \mu - \delta}{\sigma}}   , & y < \mu - \delta \\
        2(y - \mu + \delta) + \sigma e^{-\frac{y - \mu + \delta}{\sigma}} - \sigma e^{\frac{y - \mu - \delta}{\sigma}} , & \mu-\delta \leq y < \mu+\delta\\
        4\delta + \sigma e^{-\frac{y- \mu + \delta}{\sigma}} - \sigma e^{-\frac{y- \mu - \delta}{\sigma}} , & y \geq \mu + \delta \, .
    \end{cases}
\end{aligned}
$$

### The quantile function

For $\sigma>0$ and $0<u<1$, we have $Q(u)=\mu+\sigma z$, where 

$$
\begin{aligned}
    z = 
    \begin{cases}
        \log [4\tau u] - \log [e^\tau - e^{-\tau}] & \mbox{for }  0<u\leq (1-e^{-2\tau})/(4\tau) \\
        z^\ast &  \mbox{for } (1-e^{-2\tau})/(4\tau) \leq u \leq 1 - (1-e^{-2\tau})/(4\tau)\\
        - \log [4\tau (1-u)] + \log [e^{\tau} - e^{-\tau}] & \mbox{for }  1 - (1-e^{-2\tau})/(4\tau) \leq u <1,
    \end{cases}
\end{aligned}
$$ 

where $\tau=\delta/\sigma$ and $z^\ast$ is a unique solution of the equation:
$$
\begin{aligned}
    u = \frac{1}{4\tau} [ 2(z+\tau) - e^{-\tau}(e^z - e^{-z})], \quad -\tau\leq z \leq \tau. 
\end{aligned}
$$

## Session Info
```{r}
sessionInfo()
```

# References

Saah, D. K., & Kozubowski, T. J. (2025).  
*A new class of extended Laplace distributions with applications to modeling contaminated Laplace data*.  
_Journal of Computational and Applied Mathematics_.  
[https://doi.org/10.1016/j.cam.2025.116588](https://doi.org/10.1016/j.cam.2025.116588)
